The Fates of Merging Supermassive Black Holes and a Proposal for a New Class of X-Ray Sources

### 2020-01-27 paper.tex
More edits between Andreas and Charles

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[]{
Charles Zivancev$^{1}$\thanks{Contact e-mail: \href{mailto:csz2104@columbia.edu}{csz2104@columbia.edu}},
Jeremiah Ostriker$^{1}$,
Andreas H.W. K\"upper$^{2,3}$
\\
% List of institutions
$^{1}$Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY 10027, USA\\
$^{2}$QuantCo, 955 Massachusetts Avenue, Cambridge, MA 02139, USA
$^{3}$Hubble Fellow\\
}
% These dates will be filled out by the publisher
\date{Last updated 2019 Nov 09; in original form 2019 November 9}

@@

\begin{abstract}
We perform follow-up N-body simulations on some of the most massive galaxies extracted from a cosmological simulation of hierarchical structure formation with total masses in the range $10^{12} M_{\sun} < M < 3\times 10^{13} M_{\sun}$ from $4\geq z \geq 0$.  After galactic mergers, we track the dynamical evolution of the infalling black holes (BHs) around their host's central BHs.  From 11 different simulations, we find that, of the 86 infalling BHs with masses > $10^4 M_{\sun}$, 36 merge with their host's central BH, 13 are ejected from their host galaxy, and 37 are still orbiting at $z=0$.  Across all galaxies, 33 BHs are kicked to a higher orbit after close interactions with the central BH binary or multiple, after which only one of them merged with their hosts.  These orbiting BHs should be detectable by their anomalous (not LMXRB) spectra.  The X-ray luminosities of the orbiting massive BHs at z=0 are in the range $10^{27}-10^{43}$ $\mathrm{erg}~\mathrm{s}^{-1}$, with an undetectable median value $10^{32}$ $\mathrm{erg}~\mathrm{s}^{-1}$.  However, the most luminous $\sim$5\% should be detectable by existing X-ray facilities.
\end{abstract}

@@

What are possible outcomes for massive black holes during galaxy mergers?  There are three possibilities: they merge or form a binary with the central massive black hole of their new host galaxy, they get ejected through three-body interactions with an existing central black-hole binary, or they remain in orbit in the merged galaxy.  Combinations of these options are also possible with, e.g., the (spin-induced) kick at merger leading to an ejection or extended orbit of the merged black hole. We will argue in this paper that a very common outcome -- especially for lower-mass black holes -- is for the injected BHs to remain as orbiting X-ray sources in their host galaxies, waiting to be identified by current observational techniques. (check if X-1 and X-2 in NGC 1313 are candidates for such objects see https://arxiv.org/pdf/astro-ph/0608066.pdf )

But how common are encounters of massive black holes?  For galaxies, we know that minor mergers are frequent from the observed evolution of the size and mass of these systems \citep{2010ApJ...718L..73V, 2008ApJ...677L...5V, 2019MNRAS.484..595M}. However, if we assume that all galaxy mergers lead to SMBH mergers we end up overpredicting the  gravitational wave background, as constrained by observations from pulsar timing arrays (PTAs, see \citealt{2008MNRAS.390..192S, 2009MNRAS.394.2255S, 2013MNRAS.433L...1S, 2014ApJ...789..156M, 2015ApJ...799..178K, 2016APS..APRR18003T, 2018ApJ...856...42S, 2018MNRAS.479.4017I, 2018ApJ...863L..36I, 2018NatCo...9..573M}). In fact, dynamical studies of black hole encounters have shown that most encounters between two supermassive black holes will result in a stalling binary, unable to get close enough for gravitational wave emission to drive the merger (cite some papers from David Meritt here).

@@

For this study, we look at the merger history of 11 exemplary galaxies across the galaxy mass spectrum extracted from a cosmological simulation of hierarchical structure formation. We investigate how, after merging with incoming galaxies, SMBHs sink into the cores of the hosts and interact with the resident black hole. We show that gravitational interactions of multiple SMBHs are most probable in high-mass galaxies with total mass $10^{12} M_{\sun} < M < 3\times10^{13} M_{\sun}$. Galaxies with lower masses have too few mergers with SMBH hosting galaxies. Galaxies with higher masses are more extended, making dynamical friction processes less efficient and hence failing to drive SMBHs into the host galaxy core.  \citep{2012MNRAS.422.1306K} and \citep{2018MNRAS.473.3410R} performed similar N-body simulations of merging galaxies and black holes, although their haloes were from dark matter-only merger trees from the Millennium Simulation \citet{2005Natur.435..629S}, which have very different merger histories compared to the hydrodynamical simulations in this study.  Additionally, most of the halo masses in the \citep{2012MNRAS.422.1306K} simulations were of mass $M_{halo} > 10^{14} M_{\odot}$.

@@

\subsection{Overview of simulation}
Our simulations focus on elliptical galaxies with central SMBHs. The galaxies were given a dark matter background potential using the Stone-Ostriker profile (\citet{2015ApJ...806L..28S}), which is a three-parameter potential-density pair, whose quantities such as density, potential, and binding energy can be written in closed form, having a profile similar to that of a truncated isothermal sphere.  The galaxies were evolved from $4\geq z \geq 0$.  Orbiting black holes were periodically introduced into the "host" galaxy following mergers and their dynamical interaction with the background potential and central SMBH were followed, as described further below.

@@

\subsection{Merger tree Data}
The merger tree data that was used as the input to our simulations is the result of previous work by \citet{2011ApJ...741...99C, 2011ApJ...742L..33C, 2012ApJ...753...17C, 2012ApJ...748..121C, 2013ApJ...770..139C}  (known as the LAOZI simulations \citet{2014ApJ...781...38C}), as well as  \citet{2012MNRAS.425..641L}, and \citet{2015ApJ...799..178K}.  A brief synopsis of the work that led to our input data is described below.

\citet{2011ApJ...741...99C} studied the "cosmic downsizing" effect (e.g., \citet{1996AJ....112..839C}) using high-resolution large-scale hydrodynamic galaxy formation simulations.  His work produced physical parameters for galaxies such as position, velocity, total mass, stellar mass, gas mass, mean formation time, mean stellar metallicity, mean gas metallicity, star formation rate, luminosities, etc.

\citet{2012MNRAS.425..641L}, using \citet{2011ApJ...741...99C}'s work as a basis, created galactic merger trees that compared the properties of in situ and accreted stellar mass of galaxies at redshift snapshots between $4 > z > 0$.

@@

The above criteria narrowed the field of eligible galaxies down to 51.  In considering time and resources, we chose the 13 galaxies that had the largest number black holes to run the simulations.  The black holes in these galaxies accounted for nearly $84{\%}$ of all orbiting black holes across all the galaxies, so it represented a reasonable balance between completeness and efficiency.

@@

In cosmological simulations it is common to overproduce stellar mass, with a general over-efficiency in the range of 2-4 times those measured in observations (\citet{1996ApJS..105...19K}, \citet{2010MNRAS.404.1111G}, \citet{2010ApJ...725.2312O}).  One typical reason cited for this overproduction is a lack or underestimation of supernova feedback effects in the form of thermal and/or kinetic energy.  However, even when thermal energy from supernova feedback is considered, the effects may be reduced due to the energy being quickly radiated away by rapidly cooling surrounding gas (\citet{1996ApJS..105...19K}).  A second common reason for overproduction of stars in cosmological simulations is a lack of modeling for AGN feedback, which may be responsible for solving the phenomenon called the "cooling flow paradox" (\citet{2001MNRAS.321L..20F}) that suppresses star formation.

@@

\subsubsection{Galaxy background potential} \label{Galaxy background potential}
In order to model the dark matter halo distribution in our simulations, we used the Stone-Ostriker profile (\citet{2015ApJ...806L..28S}), which is a three-parameter potential-density pair.  Quantities such as density, potential, and binding energy can be written in closed form.  It is essentially an analytic form of a finite, cored isothermal mass distribution:
\begin{equation} \label{jerry}
\rho(r) = \frac{\rho_c}{(1+r^2/r_{c}^2)(1+r^2/r_{h}^2)}.
\end{equation}
Here $\rho_c$ is the central density, $r_c$ is the core radius, and $r_h$ is the outer halo radius.

In order to parameterize this profile, $r_c$ was given an initial value of 100 pc, which is reasonable for a cored massive system.  To calculate $r_h$, we began by using the stellar mass to calculate the velocity dispersion at the galaxy's effective radius.  In \citet{2014ApJ...789..156M}, the galaxy's effective radius $R_{e}$ and velocity dispersion $\sigma(R_e)$ at the effective radius are:
\begin{equation} \label{re}
R_{e} = 2.5 \mathrm{kpc}\left(\frac{M_*}{10^{11}M_{\odot}}\right)^{0.73}(1+z)^{-0.98},
\end{equation}
\begin{equation} \label{sig}
\sigma(R_{e}) = 190\mathrm{km}\;\mathrm{s}^{-1}\left(\frac{M_{*}}{10^{11}M_{\odot}}\right)^{0.2}(1+z)^{0.47},
\end{equation}

We can then equate the value for $\sigma({R_e})$ obtained from Equation \ref{sig} to the analytic expressions for $\sigma$ in the Stone-Ostriker profile (cf. Eqns 9 and A1-A4).  The only unknown is $r_h$, which we can solve for using a simple recursive Newton method.  Whether $\sigma_{near}$ or $\sigma_{far}$ is used from Stone-Ostriker is determined by whether $R_e$ is less than or greater than $\sqrt{r_c r_h}$.

The central density, $\rho_c$, can be found from Equation 5 in Stone-Ostriker:
\begin{equation} \label{rhoc}
M_\mathrm{DM} = \frac{2\pi^2r_{c}^2r_{h}^2\rho_c}{r_h+r_c}.
\end{equation}

At each iteration in the code, $r_h$ is updated according to the stellar mass and Equations \ref{re} and \ref{sig}.  The core radius, $r_c$, is recalculated only if an orbiting black hole enters within $r_c$.  The work done by diffusion, as described in Section \ref{psd} is calculated, the total potential energy and dynamical friction are updated, and $r_c$ is solved for from Equation 8 in Stone-Ostriker based on conservation of energy.

\subsubsection{Phase-space diffusion} \label{psd}
THIS SECTION TAKEN VERBATIM FROM \citet{2017MNRAS.467.4180S}, SECTION E1.  NEED TO MODIFY.
Weak encounters with background stars will let the SMBHs diffuse through phase space while they are orbiting within the gravitational potential of the galaxy. The diffusion can be expressed as change in velocity of an SMBH by $\Delta \vec{v}$ per unit time. We can split this change into a component along the direction of motion of the SMBH, and one perpendicular to that. Following \citet{2008gady.book.....B}, the diffusion coefficients can be expressed as 

@@

D[\Delta v_\parallel] & = & -\frac{4\pi G^2\rho(r)M_\bullet\ln\Lambda}{\sigma^2}f(\chi),\\
D[(\Delta v_\parallel)^2] & = & \frac{4\sqrt{2}\pi G^2\rho(r)M_\bullet\ln\Lambda}{\sigma}\frac{f(\chi)}{\chi},\\
D[(\Delta \vec{v}_\bot)^2] & = & \frac{4\sqrt{2}\pi G^2\rho(r)M_\bullet\ln\Lambda}{\sigma}\left[\frac{\mbox{erf}(\chi)-f(\chi)}{\chi}\right],

@@

where $\Delta v_\parallel \equiv \Delta \vec{v}\cdot\vec{v}/v$ is the velocity change in direction of motion, and $\Delta \vec{v}_\bot \equiv \Delta \vec{v} - \Delta v_\parallel \cdot\vec{v}/v$ is the velocity change perpendicular to the direction of motion. Here, $M_\bullet$ is the mass of the black hole, and $\chi = \frac{v}{\sqrt{2}\sigma(r)}$. The function $f(\chi)$ is given by 

@@

\Lambda \equiv \left(\frac{M_{NSC}}{M_\bullet}\right)\left(\frac{r}{r_h}\right).

@@

We can identify Equation \ref{eq:df} as the dynamical friction term. The second term introduces a variance of the friction term, and even allows the SBHs to be accelerated when the velocity of a SBH is sufficiently small. The third term introduces a change in velocity perpendicular to the direction of motion of the SBH. It is a randomly oriented vector, and hence causes the SBHs to depart from smooth orbits due to the small random velocity kicks. The last two terms will establish that the SBHs are ultimately in energy equipartition with the background stars.

@@

We compute the diffusion coefficients for each black hole at each time step, and modify its velocity on a Monte Carlo basis. For each time step we draw a random orientation before adding the perpendicular velocity change to the respective SBH. Hence, the SBH's modified velocity, $v_f$, is computed using

@@

The change of energy, $\mbox{d}E_{BH}$, of the orbiting black hole due to phase-space diffusion is given back to the stellar background potential, with $\mbox{d}E = -\mbox{d}E_{BH}$. As a consequence of this energy transfer, inspiralling black holes will cause an expansion of the stellar system. For this purpose we calculate the change in potential energy, $\mbox{d}W$, of the stellar system using

@@

where we made use of the virial theorem $2T+W =0$. With this change in potential energy we can calculate a new radius for the stellar background potential at each integration step. For the Plummer sphere the new scale radius can be calculated as
a_{new} = a\left(1+\frac{32a\,\mbox{d}W}{3\pi GM_{NSC}^2}\right)^{-1}.

\subsubsection{Gravitational wave recoils}
The code \textsc{AR-Chain} includes post-Newtonian terms up to order 2.5. The SMBHs can therefore merge via gravitational wave emission. We include gravitational wave recoils following the prescription outlined in \citet{2015ApJ...799..178K}, which is based on the fitting formula by \citet{2012PhRvD..85h4015L}.  We assume that a merger will be inevitable when the separation between two SMBHs becomes smaller than 1.0x the Schwarzschild radius. At the moment of the merger, we also assume that the spin vectors of the two SBHs are randomly aligned.

A result of the anisotropic emission of gravitational waves is that the merged binary often experiences a linear kick in the opposite direction of the emission of the GW due to conservation of linear momentum. This kick can result in velocities of the merged binary from approximately $200-5000\ km s^{-1}$ (see \citet{2007PhRvL..98w1101G}, \citet{2007PhRvL..98i1101G}, \citet{2007PhRvL..98w1102C}, \citet{2011PhRvL.107w1102L}).  In our simulations, the binaries never stray more than a few tens of parsecs from the nucleus of the host galaxies.

@@

\section{Results}\label{sec:results}
Here we review the primary results of our simulations, focusing on the various interactions of the orbiting black holes with their host SMBHs and the final luminosity characteristics of those that did not merge.

\subsection{Mergers, Ejections/Kicks and Orbiting SMBHs}

\subsubsection{Overview of Results}\label{sec:results_overview}
Figure \ref{fig:meosmbh} shows the distribution of merged, ejected and orbiting black holes across all galaxy simulations as a function of guest black hole mass.

@@

\subsubsection{Mergers}\label{sec:mergers}
The tendency of larger mass black holes to merge with their host can perhaps be more clearly seen by looking at the merger mass ratio between the incoming black hole and the host.  Figure \ref{fig:q_merge_fraction} displays the fraction of galaxy mergers that result in black hole mergers versus the merger mass ratio, defined as $m_2/m_1$, where $m_2$ is the orbiting black hole and $m_1$ is the host black hole.  The plot monotonically increases with increasing mass ratio, reaching nearly unity for mass ratio of $10^{-1}$, meaning nearly all galaxy mergers with a black hole mass ratio greater than $10^{-1}$ resulted in the orbiting black hole merging with the host.

@@

\subsubsection{Kicked, Ejected and Orbiting Black Holes}\label{sec:kicked_orbiting}
Of the 37 orbiting black holes at $z=0$, 18 of them were kicked to higher orbits during traversal of their host galaxy.  If one counts the 13 ejected black holes, then of the 50 SMBHs that had not eventually merged with their host, 32 of them were kicked out of the galactic core as a result of three-body interactions.  If they did not merge, it was not for a lack of effort.

Figure \ref{fig:kicked_stats} shows the mass distribution of kicked and not-kicked black holes, including those that merged and those that did not merge.  The vast majority of larger SMBHs ($\goa\, 10^7\, M_{\odot}$) were not kicked to higher orbits nor ejected from their galaxies, which at least partly explains why most of the merged SMBHs are in the greater mass range.  In contrast, most of the SMBHs below $10^7\, M_{\odot}$ are kicked (with a few of them being ejected).  In our simulations, no black holes that are kicked to higher orbits ever returned to merge with their central host SMBH.

Figure \ref{fig:mvr} is a realization of the final radii of the remaining wandering black holes at $z=0$.  There does not appear to be a correlation between mass and final radius.  However, it is evident from the plot that all but one of the black holes that had not been kicked by $z=0$ have mostly approached and entered within the galactic core radius, $r_c$.

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{plots/"Summary Results All Galaxies".png}
\caption{Histogram of merged, ejected, and orbiting SMBHs at z=0 as a function of BH mass.  Ejected black holes span almost the entire mass range, while lower-mass black holes are orbiting and higher mass black holes tended to merge.  Colored arrows along upper x-axis represent median masses of respective sets.  Histogram bars are not stacked.}
\label{fig:meosmbh}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{plots/kicked_stats.png}
\caption{Histogram of kicked and not-kicked SMBHs at z=0.  Colored arrows along upper x-axis represent median masses of respective sets.  Histograms are not stacked.}
\label{fig:kicked_stats}
\end{center}
\end{figure}

>>>>

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[]{
Charles Zivancev$^{1}$\thanks{Contact e-mail: \href{mailto:csz2104@columbia.edu}{csz2104@columbia.edu}},
Jeremiah Ostriker$^{1}$,
Andreas H.W. K\"upper$^{2,3}$
\\
% List of institutions
$^{1}$Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY 10027, USA\\
$^{2}$QuantCo, Inc., 955 Massachusetts Avenue, Cambridge, MA 02139, USA\\
$^{3}$Hubble Fellow\\
}
% These dates will be filled out by the publisher
\date{Last updated 2020 Jan 16; in original form 2020 Jan 16}

@@

\begin{abstract}
We perform follow-up N-body simulations on some of the most massive galaxies extracted from a cosmological simulation of hierarchical structure formation with total masses in the range $10^{12} M_{\sun} < M < 3\times 10^{13} M_{\sun}$ from $4\geq z \geq 0$.  After galactic mergers, we track the dynamical evolution of the infalling black holes (BHs) around their host's central BHs.  From 11 different simulations, we find that, of the 86 infalling BHs with masses > $10^4 M_{\sun}$, 36 merge with their host's central BH, 13 are ejected from their host galaxy, and 37 are still orbiting at $z=0$.  Across all galaxies, 33 BHs are kicked to a higher orbit after close interactions with the central BH binary or multiple, after which only one of them merged with their hosts.  These orbiting BHs should be detectable by their anomalous (not LMXRB) spectra.  The X-ray luminosities of the orbiting massive BHs at z=0 are in the range $10^{27}-10^{43}$ $\mathrm{erg}~\mathrm{s}^{-1}$, with a currently undetectable median value of $10^{33}$ $\mathrm{erg}~\mathrm{s}^{-1}$.  However, the most luminous $\sim$5\% should be detectable by existing X-ray facilities.
\end{abstract}

@@

What are possible outcomes for massive black holes during galaxy mergers?  There are three possibilities: they merge or form a binary with the central massive black hole of their new host galaxy, they get ejected through three-body interactions with an existing central black-hole binary, or they remain in orbit in the merged galaxy.  Combinations of these options are also possible with, e.g., the (spin-induced) kick at merger leading to an ejection or extended orbit of the merged black hole. We will argue in this paper that a very common outcome -- especially for lower-mass black holes -- is for the injected BHs to remain as orbiting X-ray sources in their host galaxies, waiting to be identified by current observational techniques.

%(check if X-1 and X-2 in NGC 1313 are candidates for such objects see https://arxiv.org/pdf/astro-ph/0608066.pdf )

But how common are encounters of massive black holes?  For galaxies, we know that minor mergers are frequent from the observed evolution of the size and mass of these systems \citep{2010ApJ...718L..73V, 2008ApJ...677L...5V, 2019MNRAS.484..595M}. However, if we assume that all galaxy mergers lead to SMBH mergers we end up overpredicting the  gravitational wave background, as constrained by observations from pulsar timing arrays (PTAs, see \citealt{2008MNRAS.390..192S, 2009MNRAS.394.2255S, 2013MNRAS.433L...1S, 2014ApJ...789..156M, 2015ApJ...799..178K, 2016APS..APRR18003T, 2018ApJ...856...42S, 2018MNRAS.479.4017I, 2018ApJ...863L..36I, 2018NatCo...9..573M}). In fact, dynamical studies of black hole encounters have shown that most encounters between two supermassive black holes will result in a stalling binary, unable to get close enough for gravitational wave emission to drive the merger \citep{1980Natur.287..307B, 2003AIPC..686..201M}.

@@

For this study, we look at the merger history of 11 exemplary galaxies across the galaxy mass spectrum extracted from a cosmological simulation of hierarchical structure formation. We investigate how, after merging with incoming galaxies, SMBHs sink into the cores of the hosts and interact with the resident black hole. We show that gravitational interactions of multiple SMBHs are most probable in high-mass galaxies with total mass $10^{12} M_{\sun} < M < 3\times10^{13} M_{\sun}$. Galaxies with lower masses have too few mergers with SMBH hosting galaxies. Galaxies with higher masses are more extended, making dynamical friction processes less efficient and hence failing to drive SMBHs into the host galaxy core.  \cite{2012MNRAS.422.1306K} and \cite{2018MNRAS.473.3410R} performed similar N-body simulations of merging galaxies and black holes, although their haloes were from dark matter-only merger trees from the Millennium Simulation \citep{2005Natur.435..629S}, which have very different merger histories compared to the hydrodynamical simulations in this study.  Additionally, most of the halo masses in the \cite{2012MNRAS.422.1306K} simulations were of mass $M_{halo} > 10^{14} M_{\odot}$.

@@

\subsection{Overview of simulation}
The computational challenge of this investigation is a wide range of time-scales. Galactic evolution and galactic orbits take hundreds of millions to billions of years, while orbital times of BH binaries may require times steps of years for accurate integration. We overcome this challenge by simplifying the model where possible, and by focussing on the main drivers of BH dynamics: the galactic background potential, dynamical friction, and strong BH-BH encounters.

Our simulations focus on elliptical galaxies with central SMBHs. The galaxies were given a dark matter background potential using the Stone-Ostriker profile \citep{2015ApJ...806L..28S}, which is a three-parameter potential-density pair, whose quantities such as density, potential, and binding energy can be written in closed form, having a profile similar to that of a truncated isothermal sphere.  The galaxies were evolved from $4\geq z \geq 0$.  Orbiting black holes were periodically introduced into the "host" galaxy following mergers and their dynamical interaction with the background potential and central SMBH were followed, as described further below.

@@

\subsection{Merger tree Data}
The merger tree data we used as the input to our simulations is the result of previous work by \citet{2011ApJ...741...99C, 2011ApJ...742L..33C, 2012ApJ...753...17C, 2012ApJ...748..121C, 2013ApJ...770..139C}  (known as the LAOZI simulations \citealt{2014ApJ...781...38C}), as well as  \citet{2012MNRAS.425..641L}, and \citet{2015ApJ...799..178K}.  A brief synopsis of the work that led to our input data is described below.

\citet{2011ApJ...741...99C} studied the "cosmic downsizing" effect (e.g., \citealt{1996AJ....112..839C}) using high-resolution large-scale hydrodynamic galaxy formation simulations.  His work produced physical parameters for galaxies such as position, velocity, total mass, stellar mass, gas mass, mean formation time, mean stellar metallicity, mean gas metallicity, star formation rate, luminosities, etc.

\citet{2012MNRAS.425..641L}, using \citet{2011ApJ...741...99C}'s work as a basis, created galactic merger trees that compared the properties of in-situ and accreted stellar mass of galaxies at redshift snapshots between $4 \geq z \geq 0$.

@@

The above criteria narrowed the field of eligible galaxies down to 51.  In considering time and resources, we chose the 11 galaxies that had the largest number black holes to run the simulations.  The black holes in these galaxies accounted for nearly $84{\%}$ of all orbiting black holes across all the galaxies, so it represented a reasonable balance between completeness and efficiency.

@@

In cosmological simulations, it is common to overproduce stellar mass, with a general over-efficiency in the range of 2-4 times those measured in observations \citep{1996ApJS..105...19K, 2010MNRAS.404.1111G, 2010ApJ...725.2312O}.  One cited reason for this overproduction is a lack or underestimation of supernova feedback effects in the form of thermal and/or kinetic energy.  However, even when thermal energy from supernova feedback is considered, the effects may be reduced due to the energy being quickly radiated away by rapidly cooling surrounding gas \citep{1996ApJS..105...19K}.  A second common reason for overproduction of stars in cosmological simulations is a lack of modeling for AGN feedback, which may be responsible for solving the phenomenon called the "cooling flow paradox" \citep{2001MNRAS.321L..20F} that suppresses star formation.

@@

\subsubsection{Galaxy background potential} \label{Galaxy background potential}
To model the dark matter halo distribution in our simulations, we used the Stone-Ostriker profile \citep{2015ApJ...806L..28S}, which is a three-parameter potential-density pair.  Quantities such as density, potential, and binding energy can be written in closed form, which speeds up computations.  It is essentially an analytic form of a finite, cored isothermal mass distribution:
\begin{equation} \label{jerry}
\rho(r) = \frac{\rho_c}{(1+r^2/r_{c}^2)(1+r^2/r_{h}^2)}.
\end{equation}
Here $\rho_c$ is the central density, $r_c$ is the core radius, and $r_h$ is the outer halo radius.

The central density, $\rho_c$, can be found from inverting Equation 5 for the galaxy's total mass in Stone-Ostriker:

\begin{equation} \label{rhoc}
M_{DM} = \frac{2\pi^2r_{c}^2r_{h}^2\rho_c}{r_h+r_c}.
\end{equation}

In order to parameterize $r_c$ and $r_h$, we can utilize the relationship between the distribution of light and stellar velocity dispersion at the effective radius, and the shape of the dark matter profile.  The effective radius $R_{e}$ and velocity dispersion at the effective radius $\sigma(R_e)$, respectively, can be obtained from (\citet{2009ApJ...706L..86N}, \citet{2010ApJ...725.2312O}, \citet{2014ApJ...789..156M}):

\begin{equation} \label{re}
R_{e} = 2.5 \mathrm{kpc}\left(\frac{M_*}{10^{11}M_{\odot}}\right)^{0.73}(1+z)^{-0.98},
\end{equation}
\begin{equation} \label{sig}
\sigma(R_{e}) = 190\mathrm{km}\;\mathrm{s}^{-1}\left(\frac{M_{*}}{10^{11}M_{\odot}}\right)^{0.2}(1+z)^{0.47},
\end{equation}

We also give $r_c$ an initial value of 100 pc, which is reasonable for a cored massive system.  We can then equate the value for $\sigma({R_e})$ obtained from Equation \ref{sig} to the analytic expressions for $\sigma$ in the Stone-Ostriker profile (cf. Eqns 9 and A1-A4).  The only unknown is $r_h$, which we can solve for using a simple recursive Newton method.  Whether $\sigma_{near}$ or $\sigma_{far}$ is used from Stone-Ostriker is determined by whether $R_e$ is less than or greater than $\sqrt{r_c r_h}$.

At each timestep in the code, $r_h$ is updated according to the stellar mass and Equations \ref{re} and \ref{sig}.  The core radius, $r_c$, is recalculated only if an orbiting black hole enters within $r_c$.  The work done by diffusion, as described in Section \ref{psd} is calculated, the total potential energy and dynamical friction are updated, and $r_c$ is solved for from Equation 8 in Stone-Ostriker based on conservation of energy.

\subsubsection{Phase-space diffusion} \label{psd}
Much of this subsection was borrowed from \citet{2017MNRAS.467.4180S}, Section E1, which made use the same modified version of \textsc{AR-Chain}.

Weak encounters with background stars and dark-matter particles will let the SMBHs diffuse through phase space while they are orbiting within the gravitational potential of the galaxy. The diffusion can be expressed as change in velocity of an SMBH by $\Delta \vec{v}$ per unit time. We can split this change into a component along the direction of motion of the SMBH, and one perpendicular to that. Following \citet{2008gady.book.....B}, the diffusion coefficients can be expressed as 

@@

D[\Delta v_\parallel] & = & -\frac{4\pi G^2\rho(r)M_{BH}\ln\Lambda}{\sigma^2}f(\chi),\\
D[(\Delta v_\parallel)^2] & = & \frac{4\sqrt{2}\pi G^2\rho(r)M_{BH}\ln\Lambda}{\sigma}\frac{f(\chi)}{\chi},\\
D[(\Delta \vec{v}_\bot)^2] & = & \frac{4\sqrt{2}\pi G^2\rho(r)M_{BH}\ln\Lambda}{\sigma}\left[\frac{\mbox{erf}(\chi)-f(\chi)}{\chi}\right],

@@

where $\Delta v_\parallel \equiv \Delta \vec{v}\cdot\vec{v}/v$ is the velocity change in direction of motion, and $\Delta \vec{v}_\bot \equiv \Delta \vec{v} - \Delta v_\parallel \cdot\vec{v}/v$ is the velocity change perpendicular to the direction of motion. Here, $M_{BH}$ is the mass of the black hole, and $\chi = \frac{v}{\sqrt{2}\sigma(r)}$. The function $f(\chi)$ is given by 

@@

\Lambda \equiv \left(\frac{M_{M_{DM}}}{M_{BH}}\right)\left(\frac{r}{r_h}\right).

@@

We can identify Equation \ref{eq:df} as the dynamical friction term. The second term introduces a variance of the friction term, and even allows the BHs to be accelerated when the velocity of a BH is sufficiently small. The third term introduces a change in velocity perpendicular to the direction of motion of the BH. It is a randomly oriented vector, and hence causes the BHs to depart from smooth orbits due to the small random velocity kicks. The last two terms will establish that the BHs are ultimately in energy equipartition with the background stars.

@@

We compute the diffusion coefficients for each black hole at each time step, and modify its velocity on a Monte Carlo basis. For each time step we draw a random orientation before adding the perpendicular velocity change to the respective BH. Hence, the BH's modified velocity, $v_f$, is computed using

@@

The change of energy, $\mbox{d}E_{BH}$, of the orbiting black hole due to phase-space diffusion is given back to the galactic background potential, with $\mbox{d}E = -\mbox{d}E_{BH}$. As a consequence of this energy transfer, inspiralling black holes will cause an expansion of the galactic mass distribution. For this purpose we calculate the change in potential energy, $\mbox{d}W$, of the host galaxy using

@@

where we made use of the virial theorem $2T+W =0$. With this change in potential energy we can calculate a new radius for the galactic background potential at each integration step. For simplicity, we assumed that the relatively low mass of the infalling black holes compared to the core mass of the host galaxies would only affect the size of the core radius, and hence ignored the feedback effect when the black holes were outside of the core radius.

\subsubsection{Gravitational wave recoils and escape}
The code \textsc{AR-Chain} includes post-Newtonian terms up to order 2.5. The SMBHs can therefore merge via gravitational wave emission. We include gravitational wave recoils following the prescription outlined in \citet{2015ApJ...799..178K}, which is based on the fitting formula by \citet{2012PhRvD..85h4015L}.  We assume that a merger will be inevitable when the separation between two SMBHs becomes smaller than 1.0x the Schwarzschild radius. At the moment of the merger, we also assume that the spin vectors of the two SBHs are randomly aligned.

A result of the anisotropic emission of gravitational waves is that the merged binary often experiences a linear kick in the opposite direction of the emission of the GW due to conservation of linear momentum. This kick can result in velocities of the merged binary from approximately $200-5000\ km s^{-1}$ (see \citealt{2007PhRvL..98w1101G, 2007PhRvL..98i1101G, 2007PhRvL..98w1102C, 2011PhRvL.107w1102L}.  In our simulations, the binaries never stray more than a few tens of parsecs from the nucleus of the host galaxies.

@@

\section{Simulation Results}\label{sec:results}\label{sec:results_overview}
Here we review the primary results of our simulations, focusing on the various interactions of the orbiting black holes with their host SMBHs. In the next section, we will investigate the luminosity characteristics of the orbiting black holes that did not merge at $z=0$.

Figure \ref{fig:meosmbh} shows the distribution of merged, ejected and orbiting black holes (stacked) across all galaxy simulations as a function of the mass of the infalling black holes.

@@

\subsection{Mergers}\label{sec:mergers}
The tendency of larger mass black holes to merge with their host can perhaps be more clearly seen by looking at the merger mass ratio between the incoming black hole and the host.  Figure \ref{fig:q_merge_fraction} displays the fraction of galaxy mergers that result in black hole mergers versus the merger mass ratio, defined as $m_2/m_1$, where $m_2$ is the orbiting black hole and $m_1$ is the host black hole.  The plot monotonically increases with increasing mass ratio, reaching nearly unity for mass ratio of $10^{-1}$, meaning nearly all galaxy mergers with a black hole mass ratio greater than $10^{-1}$ resulted in the orbiting black hole merging with the host.

@@

Approximately 50\% of the mergers appeared to have been directly assisted by a hard three-body encounter with a third rendezvousing black hole, which as a consequence got kicked to a higher orbit (and in a few instances ejected).  A large fraction of the remaining BH mergers were accelerated through soft encounters with infalling black holes. Our simulations suggest that few-body encounters are in fact a significant driver of BH-BH mergers in galactic centers.

\subsection{Kicked, Ejected and Orbiting Black Holes}\label{sec:kicked_orbiting}
Of the 37 orbiting black holes at $z=0$, 18 of them were kicked to higher orbits during traversal of their host galaxy.  If one counts the 13 ejected black holes, then of the 50 SMBHs that had not eventually merged with their host, 32 of them were kicked out of the galactic core as a result of three-body interactions.  If they did not merge, it was not for a lack of effort.

Figure \ref{fig:kicked_stats} shows the mass distribution (stacked) of kicked and not-kicked black holes, including those that merged and those that did not merge.  The vast majority of larger SMBHs ($\goa\, 10^7\, M_{\odot}$) were not kicked to higher orbits nor ejected from their galaxies, which at least partly explains why most of the merged SMBHs are in the greater mass range.  In contrast, most of the SMBHs below $10^7\, M_{\odot}$ are kicked (with a few of them being ejected).  In our simulations, no black holes that are kicked to higher orbits ever returned to merge with their central host SMBH.

Figure \ref{fig:mvr} is a realization of the final radii of the remaining wandering black holes at $z=0$.  There is a slight inverse relationship between black hole mass and final radius (solid line), but with a wide dispersion.  It is also evident from the plot that all but two of the black holes that had not been kicked by $z=0$ have mostly approached and entered within the galactic core radius, $r_c$ (which never exceeded 1 kpc in any of the galaxies).

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{plots/"Summary Results All Galaxies".png}
\caption{Histogram of merged, ejected, and orbiting SMBHs at z=0 as a function of BH mass.  Ejected black holes span almost the entire mass range, while lower-mass black holes are orbiting and higher mass black holes tended to merge.  Colored arrows along upper x-axis represent median masses of respective sets.  Histogram bars are stacked.}
\label{fig:meosmbh}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{plots/kicked_stats.png}
\caption{Histogram of kicked and not-kicked SMBHs at z=0.  Colored arrows along upper x-axis represent median masses of respective sets.  Histograms are stacked.}
\label{fig:kicked_stats}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{plots/"Final Radius v Mass of Orbiting BHs".png}
\caption{Final Radii of remaining orbital black holes at z=0.  Kicked black holes generally had a low likelihood of finding their way back to the center by z=0.  There is a slight inverse relationship between BH mass and final distance from center (solid line).}
\label{fig:mvr}
\end{center}
\end{figure}
